Fly away Peter, come back Paul: the gains and losses of avian Non-coding RNAs

### 2013-09-23 paper/manuscript.tex
updated manuscript.

\title{Non-coding RNAs of bird genomes}

@@

\section*{Results}
In the following we explore the conservation patterns of the major
classes of ncRNA in further detail. The collection of ncRNA sequences
is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. Consequently the bulk of the
families discussed below are either conserved across the vertebrates
or have been apparently lost or diversified in the avian
lineage(s).

\subsection*{Classic RNAs: LUCA and LECA}

Many RNA families constitute the most evolutionarily conserved genes
across all life on this planet \cite{Jeffares:1998}. Examples of RNAs
derived from the Last Universal Common Ancestor (LUCA) include, the
transfer RNAs (tRNA), ribosomal RNAs (rRNA), RNA components of RNase P
(RNase P RNA), RNase MRP (RNase MRP RNA) and the signal recognition
particle (SRP RNA). Other classes of RNA are likely to have been
components of the Last Eukaryotic Common Ancestor (LECA) include the
telomerase RNA, major spliceosomal RNAs (U1, U2, U4, U5, and U6) and
the minor spliceosomal RNAs (U11, U12, U4atac, and U6atac)
\cite{Hoeppner:2012}.

%\subsubsection*{Losses}

Unsurprisingly, the bulk of these classes of RNAs are well represented
across the bird genomes (See Figure~\ref{fig:1}). However, there
appear to have been ``losses'' of a few of these RNAs in certain bird
species. Some of these may be due to sequence divergence, of which
there are several notable examples e.g.
\cite{Leonardi:2008,Webb:2008,Mao:,Lai:2010,Chan:2011}. Other losses
may be due to incomplete genome coverage.

A number of the classic RNAs are incorporated into RNA-protein
complexes (RNPs) involved in core cellular processes. An example of
this are the spliceosomal RNAs. Based upon the presence/absence
patterns of the major spliceosomal RNAs they are all well represented
in these genome sequences. Except for the U4 RNA in cormorant and the
U5 RNA in the bee eater. Again these two genomes are low coverage
suggesting these genes weren't captured in the current assembly. The
minor spliceosomal RNAs are more interesting, the U4atac and U11
snRNAs show widespread patterns of loss, even in some of the high
coverage genomes. These RNAs are frequently missed in bioinformatic
screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
sequences that have diverged beyond the ability of detection by
covariance models \cite{Marz:2008}.

The telomerase RNA is also largely missing from the avian
annotations. This RNA acts as a template for the telomerase enzyme
that extends the telomeres found on chromosome ends. It is only found
in the chicken, bald eagle, kea, budgerigar, crow and
zebrafinch. Homology searches searches with the telomerase reverse
transcriptase (TERT) protein show that the protein component of the
telomerase RNP is conserved across all the bird genomes (data not
shown). This pattern of presumably divergent telomerase RNA and
conserved telomerase protein has been noted previously, most notably
in the fungi \cite{Leonardi:2008,Webb:2008}.

The the RNA components of RNase P and RNase MRP also appear to have
undergone dramatic losses within the bird lineage. RNase P is required
for the maturation of tRNA, the paralogous enzyme, RNase MRP is
required for the maturation of rRNA. Each RNP cleaves smaller RNAs
from larger transcripts \cite{Lopez:2009}. It is unlikely that the
these genes have been lost in any of the birds. Homology searches with
the RNase associated protein coding genes (POP1, POP4, POP5, POP7,
RPP1, RPP14, RPP25, RPP38, RPP40 and RPR2), identified viable homologs
of each in all of the bird genomes \cite{Rosenblad:2006} (data not
shown). This suggests that the bird RNase P and MRP RNAs may have
diverged slightly from the canonical models.

The 5.8 S component of the ribosome in the turtle, turkey bustard,
hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
bee eater and falcon appears to have been lost (See
Figure~\ref{fig:1}). The rRNA repeats are frequently not assembled,
consequently it is not surprising to see ``losses'' in these \cite{Floutsakou:2013}.
Furthermore, the genomes for these species are also
low-coverage.

\subsection*{Small nucleolar RNAs}

Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
in the maturation of other functional RNAs \cite{Gardner:2010}. The
bulk of the characterised snoRNAs guide either methylation or
pseudouridylation modifications, primarily of rRNAs but also
spliceosomal RNAs. The two types of modifications are guided by two
different types of RNA, the box C/D and the H/ACA snoRNAs
respectively, each with a characteristic cohort of motifs and
secondary structures \cite{Marz:2011}.

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.
Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 locii of the ancestral
vertebrate genome. We investigated these losses further.

The first cluster is found at chr11:62620797-62622484 on the human
genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
human-yeast conserved snoRNAs. These snoRNAs are located in the
inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
alligator and turtle genomes within a 3-4 KB locus, yet these have
largely been lost in the birds. Notable exceptions are five of the
eight are located in the tinamou genome, these are located on the same
scaffold and are within 2 KB of each other. Implying that SNHG1 is
conserved in the tinamou. Loci with four of the eight snoRNAs can be
found in zebrafinch, ground-finch, and bald eagle. Still, three of the
eight are located in the ostrich, crow, and cockoo genomes, again
within 2 KB of each other on the same scaffolds. This complex pattern
of loss could be attributed to many different models, e.g. multiple
losses in birds, poor homology modelling or incomplete genome
sequences.

%Furthermore our results contradict the text in the paper. We find
%SNORD26 stated to be lost in sauropsids in the 5 species:
%C.vociferus, M.unicolor, H.leucocephalus, L.discolor and T.guttata.

%%GAS5?
%SNORD81, SNORD47, SNORD80, SNORD79, SNORD78, SNORD44, SNORD77, SNORD76, SNORD75, SNORD74
%egrep 'SNORD81;|SNORD47;|SNORD80;|SNORD79;|SNORD78;|SNORD44;|SNORD77;|SNORD76;|SNORD75;|SNORD74' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

The second cluster is located at chr19:49993222-49994231 on the human
genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
within a 1 KB genomic region. The turtle and alligator genomes retain
the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
the same scaffold. Within the bird genomes, the crow and rifleman each
retain a single SNORD33 and SNORD34 gene on the same scaffold. While
the ground-finch and bald eagle retain a single SNORD33 and the
zebrafinch and seriema retain a single SNORD34 (See
Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
gene, ribosomal protein L13a (RPL13A). Based on BLASTP (version
2.2.18) homology searches for the RPL13A gene, the protein is
conserved in the human and turtle genomes and in the bald eagle, crow,
rifleman and zebrafinch avian genomes (data not shown). Therefore the
conservation patterns of the RPL13A host gene and corresponding
intronic snoRNAs largely mimic each other, each independently supports
a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
hosts in the bird genomes.

\subsection*{MicroRNAs}
MicroRNAs are an important class of non-coding RNA. They have been
found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
\cite{Hinas:2007,Avesson:2012}, Viridiplantae
\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}. The
miRNAs have been shown to regulate the expression of large numbers of
messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
22 nucleotides long which is usually processed from a larger RNA that
is characterised by a stable hairpin-shaped secondary structure.

Chicken and zebrafinch are the only birds with previously annotated
microRNAs. We searched for homologs of these and other vertebrate
microRNAs in the genomes of the 48 birds, American alligator and green
turtle. Overall, we annotate a total of 16617 putative microRNA loci,
homologous to 543 known microRNA genes, of which 487 are annotated in
chicken and/or zebra finch, while 56 have been so far known only in
non-avian vertebrates. The numbers of annotated loci in the individual
species are approximately equal - 300-400 per species, except for the
turkey (Meleagris gallopavo) where we identified 543 sequences
homologous to known microRNAs.

Nearly all the microRNAs that are broadly conserved in fish,
amphibians and mammals are also conserved in the birds. Nevertheless,
there are obvious instances of microRNAs lost in all birds. For
example, mammalian, amphibian and fish genomes contain three loci of
clustered microRNAs from the mir-17 and mir-92 families. One of these
clusters (mir-93~25) was not found in turtles, crocodiles and
birds. In addition, we can confidently identify a further 3 microRNA
families that are present in mammals, and turtle and/or crocodile, but
not in any avian genome (mir-150, mir-208, mir-590). This suggests
that these sequences were lost in the last common ancestor of
archosaurs or birds. There are also a number of microRNAs that are
predicted to be present in turtles and/or crocodiles, and only a small
number of bird genomes. Indeed, there are many missing annotations,
species-specific and otherwise, that are not consistent with the
consensus phylogeny, and could be due to either incomplete genomes or
widespread microRNA loss.

The turkey genome contains a high number (190) of microRNAs so far
found only in chicken, which account for the higher number of
annotated sequences in this genome compared with other birds. This is
consistent with its phylogenetic position as the closest chicken
relative among the examined birds. However, 101 chicken microRNAs
have no homolog in the turkey or other bird genomes, suggesting that
these genes are chicken-specific. This is consistent with previous
reports of large number of speciesspecific microRNAs in all animals,
and supports the view of fast microRNA turnover during animal
evolution.

\subsection*{Long non-coding RNAs}

%%{\bf MAY BE MOVED INTO THE lncRNA PAPER!}

As mentioned in the introduction, lncRNAs are a diverse group of RNAs
that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have mainly been
characterised in mammalian species, particularly human and mouse.

In general Rfam cannot include the entire length of any large, spliced
RNAs. This is a limitation of the covariance-models used for the
homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
short, well-conserved regions with evolutionarily conserved secondary
structures are included in Rfam, we will refer to these as RNA-domains
\cite{Burge:2013}.

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains appear to be consistently
preserved in the birds. The annotations of these domains are all in
the same genomic region supporting a high degree of evolutionary
conservation for the entire lncRNA. In particular the HOXA11-AS1,
PCA3, RMST, Six3os1, SOX2OT and ST7-OT3 lncRNAs have multiple well
conserved RNA-domains (See Figure~\ref{fig:5}).

@@

The RMST (Rhabdomyosarcoma 2 associated transcript) RNA-domains 6, 7,
8, and 9 are conserved across the birds. In each bird the gene order
was also consistent with the human ordering. In the alligator and
turtle an additional RNA-domain was predicted in each, these were
RNA-domains 2 and 4 respectively, again the ordering of the domains
was consistent with human. This suggests that the RMAST lncRNA is
highly conserved. However, little is known about the function of this
RNA. It was originally identified in a screen for differentially
expressed genes in two Rhabdomyosarcoma tumor types \cite{Chan:2002}.

In addition, the lncRNA DLEU2, is well conserved conserved across the
vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,
both of which are also well conserved across the vertebrates (See
Figure~\ref{fig:3}). DLEU2 is thought to be a tumor-suppressor
gene as it is frequently deleted in malignant tumours \cite{Lerner:2009,Klein:2010}.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad ranch of tissues. Extensive
research on BRCA1 has shown that it is involved in DNA repair
\cite{}. The function of NBR2 remains unknown, yet its conservation
across the vertebrates certainly implies a function (See
Figure~\ref{fig:5}).

@@

\subsection*{Cis-regulatory elements}
%% Histone3, IRE, IRES, K chan RES, SECIS, Antizyme FSE, Vimentin3, GABA3,
%% CAESAR,

The cis-regulatory RNAs are a group of RNA structures encoded on
mRNAs. Generally they are involved in regulating the expression of the
mRNA they are encoded within. Others may recode the translated protein
product into an alternate sequence.

This group includes the iron response element (IRE) \cite{Stevens:}
and the histone $3^\prime$ UTR (histone3)
\cite{Davila_Lopez:2008a}. These are structured motifs bound by
regulatory proteins. The selenocysteine insertion sequence (SECIS) is
a structured motif that recodes UGA stop codons to selenocysteines
\cite{Lambert:2002} and the GABRA3 stem-loop is a structure recognised
by the ADAR enzyme family. This enzyme edits adenine nucleotides to
inosine, in this case recoding an isoleucine codon to methionine in
exon 9 of the GABRA3 gene \cite{Ohlson:2007}.

These regulatory elements and others, including an internal ribosome
entry site (IRES), potassium channel RNA editing signal (K chan RES),
Antizyme RNA frameshifting stimulation element (Antizyme FSE),
vimentin $3^\prime$ UTR protein-binding region (Vimentin3) and a
connective tissue growth factor (CTGF) $3^\prime$ UTR element (CAESAR)
are conserved across a diverse group of vertebrates, including the
bird lineages explored here (See Figure~\ref{fig:1}).

\subsection*{Exceptional RNAs}

%Background to this group...

A number of other ncRNAs can also be found in Eukaryotic
genomes. These do not fit into the main classes of RNA but still
perform vital roles in the function and evolution of Eukaryotes.
Their functions are diverse and many have not yet been characterised.

An example of an uncharacterised RNA is the ultraconserved element,
uc.338 (also known as TUC338)
\cite{Bejerano:2004,Bejerano:2006,Braconi:2011}. The uc.338 element
is derived from a short interspersed element (SINE) called the
lobe-finned fishes SINE (LF-SINE) as it is conserved between the
coelacanth and mammals \cite{Bejerano:2006}. Analysis of the
expression of uc.338 implies that it plays a role in the progression
of hepatocellular carcinoma, possibly by influencing cell growth
\cite{Braconi:2011}. This RNA is conserved in the birds and appears to
have been duplicated in several lineages.

The Y RNA is an enigmatic ncRNA where we know very little about the
function. It was discovered in the 1980s in ribonucleoprotein
complexes \cite{Lerner:1981}. The function of the Y RNAs remain
unknown, but evidence is emerging that they may be associated with DNA
replication \cite{Christov:2006}. There are 4 functional Y RNAs
encoded in the human genome, Y1, Y3, Y4 and Y5. However, there are
hundreds of pseudogenised copies of the Y RNA scattered throughout the
human genome \cite{Mosig:2007}. In the birds and other lizards, we
identify between two and seven Y RNA paralogs (See
Figure~\ref{fig:1}).

The Vault RNA forms a major component of the vault ribonucleoprotein
complex, this is one of the largest particles found in the vertebrate
cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
not much is known about the function of Vault. The Vault RNA has been
shown to be broadly conserved in metazoans
\cite{Stadler:2009}. However, in the bird lineages it appears to have
either been lost or diversified.

% Vault, 7SK, HAR1A

%% \subsection*{Pseudogenes}

%% The human genome is littered with pseudogenised copies of ncRNAs. The
%% Dfam database of human repeats estimates that approximately $54\%$ of
%% the human genome is composed of repeat elements
%% \cite{Wheeler:2013}. Of these annotations XXX\% are derived from
%% ncRNAs, including the large Alu family of repeats, which is derived
%% from a retransposed SRP RNA pseudogene \cite{}.

%% tRNA, RSV RNA,
%%  \cite{International_Chicken_Genome_Sequencing_Consortium:2004}

%% \emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB

\subsection*{Contamination}

Bacterial families can be used to identify problematic sequences that
are likely to be the result of contamination from non-avian sources.

@@

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. All
matches above the curated GA threshold were included. Subsequently,
all hits with an E value greater than 0.0005 were discarded, so only
matches which passed the model-specific GA threshold and had an E-value smaller than 0.0005 were retained.

tRNA-scan version 1.3.1, ... \cite{Lowe:1997,Chan:2009}

Rfam matches and the tRNA-scan results for families belonging to the
same clan were then competed so that only the best match was retained.
Matches were further sorted into taxonomic groups {\bf ONLY MATCHES IN 5 OR MORE ...}

@@

\begin{figure}[ht]
  \includegraphics[width=0.4\textwidth]{figures/RNA.pdf}
  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
  \includegraphics[width=0.4\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
  \caption[]{This figure illustrates the number of copies of Rfam and
    tRNAscan annotations for each RNA family in each avian species and
    outgroups. The families are ordered alphabetically vertically and
    the species are in phylogenetic order horizontally. Darker shades
    indicate high copy numbers, lighter shades indicate fewer, white
    indicates no predictions were made.}\label{fig:1}
\end{figure}

%% \begin{figure}[ht]
%% \includegraphics[width=0.65\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
%% \caption[]{.}\label{fig:2}
%% \end{figure}
\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
    the left.}\label{fig:3}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf lncRNA} families on the right and {\bf tRNA}
    families on the left.}\label{fig:5}
\end{figure}

>>>>

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Fly away Peter, come back Paul: the gains and losses of avian Non-coding RNAs}

@@

\section*{Results}
In the following we explore the conservation patterns of the major
classes of ncRNA in further detail. The collection of ncRNA sequences
is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. We focus our analysis on the
unusual results within the avian lineages. These are either
unexpectedly well-conserved RNAs or unexpectedly poorly-conserved
RNAs. The former are RNAs we would not have expected to be conserved
between the birds and the organisms these genes were initially
identified in; Usually, this hypothesis is based upon the function of
the RNA which is not conserved in avian species. The latter are
apparent losses of RNA genes that were expected to be conserved;
Usually, this hypothesis is based upon the conservation of these RNAs
in other vertebrate species. We have three models of ``loss'' that we
use to explain the data: Firstly, these could be genuine cases of an
ancestral gene-loss along the avian lineage. Secondly, this could be a
case of ``divergence'' where a RNA gene has undergone significant
sequence and structural alterations, so much so, that our homology
search tools can no longer detect a relationship between vertebrate
exemplars and the avian varieties. Thirdly, we consider the
possibility that the available genome assemblies have independently
failed to capture these genes.

\subsection*{Unusually, well conserved RNAs}

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.

In general Rfam cannot include the entire length of any large, spliced
RNAs. This is a limitation of the covariance-models used for the
homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
short, well-conserved regions with evolutionarily conserved secondary
structures are included in Rfam. By analogy to protein-domains, we
refer to these as RNA-domains \cite{Burge:2013}.

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus
supporting a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:5}).

@@

The RMST (Rhabdomyosarcoma 2 associated transcript) RNA-domains 6, 7,
8, and 9 are conserved across the birds. In each bird the gene order
was also consistent with the human ordering. In the alligator and
turtle an additional RNA-domain was predicted in each, these were
RNA-domains 2 and 4 respectively, again the ordering of the domains
was consistent with human. This suggests that the RMST lncRNA is
highly conserved. However, little is known about the function of this
RNA. It was originally identified in a screen for differentially
expressed genes in two Rhabdomyosarcoma tumor types \cite{Chan:2002}.

In addition, the lncRNA DLEU2, is well conserved conserved across the
vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,
both of which are also well conserved across the vertebrates (See
Supplemental Figure~2). DLEU2 is thought to be a tumor-suppressor gene
as it is frequently deleted in malignant tumours
\cite{Lerner:2009,Klein:2010}.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad ranch of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:5}).

{\bf more detail on the prostate lncRNAs?}

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA SNORD93 which we show has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

@@

%% Also, it would be good to check regarding the HOTAIRM/HOTTIP/HOXA11AS
%% etc lncRNA elements that their syntenic location relativ to the HOXA genes
%% (similar for SOXAS...) is correct. Mario checked for HOTAIRM and HOXA1/HOXA2
%% and this is fine, it would be good for my piece of mind though it is
%% at least OK in chicken, turkey, duck, zebrafinch. Also a word of caution
%% of the Rfam models: if you look at HOTAIRM1_2, there's basically no
%% covariation, looks pretty much like an ultraconserved element, so
%% technically we don't really know whether this is an RNA element or,
%% say, a TF binding site cluster that is functional at DNA level. Hence any
%% evidence that we have indeed a transcript would be nice to have. We
%% checked for HOTAIRM1 and it looks good in chicken. For the rest, it would be
%% nice to have seen transcription it at least one bird :)

\subsection*{RNA Losses, divergence or missing data?}

Much of the number of apparent losses and reduction in genomic
sequence has been extensively discussed elsewhere
\cite{Organ:2007}. Unsurprisingly, this reduction is reflected in the
copy-number of RNA genes. Some of the most dramatic examples are the
transfer RNAs and pseudogenes which average $\approx 900$ and $\approx
580$ copies in the human, turtle and alligator genomes, the average
copies numbers of these drop to $\approx 280$ and $\approx 100$ copies
in the avian genomes.

The absence of several well-conserved ncRNA families from many or even most
bird genomes is unlikely to represent true gene losses. This concerns in
particular the telomerase RNA, the RNA components of RNase P and MRP, the
minor spliceosomal snRNAs U4atac and U11, the selenocystein tRNA (tRNA-Sec)
as well as the vault RNAs. In order to get an idea to what extent the
absence of these RNAs from the \texttt{infernal-based} annotation is caused 
by sequence divergence beyond the thresholds of the Rfam CMs and/or 
missing or incomplete data, we complemented our analysis by dedicated
searches for a few of these RNA groups.

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned for
specificity and thus misses several occurrances that are easily found by
\texttt{blastn} with $E\le 10^{-30}$. In some cases the sequences appear
degraded at the ends, which may be explained e.g.\ by low sequence quality
at the very ends of contigs or scaffolds. A \texttt{blastn} search also
readily retrieves additional RNAse P and RNAse MRP RNAs, capturing only the
best conserved regions. In many cases these additional candidates are
incomplete or contain undetermined sequence, explaining why they are missed
by the CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs
in the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very poor
sequence conservation and rapid variations in size that make it notoriously
hard to identify by homology search \cite{Xie:08a}. The poor return thus
does not come as a surprise. Since \texttt{blastn} searches remained
unsuccessfull we constructed a sauropsid-specific CM for the vault RNA. In
addition to the hits identified by the Rfam model we obtained three
additional homologs. Vault RNAs, with a size of about 100 nt, exhibit
conserved sequence patterns only at their ends, with essentially
unconstrained sequence in the central part. Their identification is also a
well-know difficult problem for homology search
\cite{Stadler:09b}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Supplemental Figure~5)
strongly suggests that that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

\subsection*{Y RNA clusters}

Vertrebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\emph{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within bird, we find an the conserved
Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most cases
consistent with breaks in the available sequence assemblies. In
several genomes we observe one or a few additional Y RNA homologs
unlinked to the canonical Y RNA cluster. These sequences can be
identified unambiguously as derived members of one of the three
ancestral paralog groups, they almost always fit less well to the
consensus (as measured by the CM bit score of paralog group specific
covariance models) than the paralog linked to cluster, and there is no
indication that any of these additional copies is evolutionarily
conserved over longer time scales. We therefore suggest that most or
all of these interspersed copies are in fact pseudogenes.

\subsection*{MicroRNA Clusters}

Nearly all microRNAs that are broadly conserved in fish, amphibians and
mammals are also conserved in the birds. Nevertheless, there are obvious
instances of microRNAs lost in all birds. For example, mammalian and
ambhibian genomes contain three loci of clustered microRNAs from the mir-17
and mir-92 families \cite{Tanzer:04}. One of these clusters (cluster II,
with families mir-106b, mir-93 and mir-25) was not found in turtles,
crocodiles and birds, see Supplemental Figure~6.

The microRNA family let-7 is the most abundant microRNA family with 14
paralogs in human. These genes also localize in 7 genomic clusters,
together with mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
we observed that cluster A - that is strongly conserved in vertebrates has
been completely lost in the avian lineage. Another obious lost in birds is
cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
hand has been retained in all oviparous animals and completely lost later,
after the split of Theria. See Supplemental Figure~7 for details.

@@

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0 \cite{Gardner:2011a,Burge:2013}. All matches above the curated GA
threshold were included. Subsequently, all hits with an E-value
greater than 0.0005 were discarded, so only matches which passed the
model-specific GA threshold and had an E-value smaller than 0.0005
were retained.

In order to obtain good annotations of tRNA genes we also ran the
specialist tRNA-scan version 1.3.1 annotation tool. This method also
uses covariance models to identify tRNAs. However it also uses some
heuristics to increase the search-speed, annotates the Isoacceptor
Type of each prediction and uses sequence analysis to infer if
predictions are likely to be functional or tRNA-derived pseudogenes
\cite{Lowe:1997,Chan:2009}.

Rfam matches and the tRNA-scan results for families belonging to the
same clan were then ``competed'' so that only the best match was
retained for any genomic region \cite{Gardner:2011a}.  To further
increase the specificity of our annotations we filtered out families
that were indentified in four or fewer of the 51 vertebrate species we
have analysed in this work. These filtered families largely
corresponded to bacterial contamination within the genomic sequences.

@@

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf lncRNA} families.}\label{fig:5}
\end{figure}
